---
title: "1.4: Stan - A probabilistic programming language"
author: "Benjamin Rosenbaum"
date: "April 1, 2019"
output:  ioslides_presentation
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```


## MCMC Samplers 

> - BUGS (WinBUGS, OpenBUGS)

> - JAGS [mcmc-jags.sourceforge.net](http://mcmc-jags.sourceforge.net/)

> - Stan [mc-stan.org](https://mc-stan.org/)

> - greta [greta-stats.org](https://greta-stats.org/)

## MCMC Samplers 

> - Stan, JAGS can be called from R (via packages)
> - model description: priors, likelihood
> - no explicit code for likelihood required
> - just tell the program how the residuals are distributed
> - takes care of MCMC sampling

## Linear regression model

$y_i \sim \text{normal}(\mu_i,\sigma)\quad i=1,...,n$

$\mu_i = a+b\cdot x_i$

- data: $x_i, y_i,\quad i=1,...,n$
- parameters: $a,b,\sigma$
- deterministic part: $\mu_i = a+b\cdot x_i$
- stochastic part: $y_i \sim \text{normal}(\mu_i,\sigma)$  

## Stan code

```{r stan_code, echo=TRUE}
stan_code = '
data {
  int n;
  vector[n] x;
  vector[n] y;
}
parameters {
  real a;  // intercept
  real b;  // slope
  real<lower=0> sigma;  // standard deviation
}
model {
  // priors
  a ~ normal(0, 10);
  b ~ normal(0, 10);
  sigma ~ normal(0, 10);
  // likelihood
  y ~ normal(a+b*x, sigma);
}
'
```

## Data block {.build}

```{r, echo=TRUE, eval=FALSE}
data {
  int n;
  vector[n] x;
  vector[n] y;
}
```

- each variable must be specified!  
- datatypes:    
  `int x;` integer (single value)      
  `real x;` continuous (single value)        
  `vector[n] x;` vector of length n       
  (see Stan documentation for more: matrices, arrays, ...)    
- each command ends with a semicolon `;`


## Parameters block {.build}

```{r, echo=TRUE, eval=FALSE}
parameters {
  real a;  // intercept
  real b;  // slope
  real<lower=0> sigma;  // standard deviation
}
```

- `//` used for comments: not read as code
- specify boundaries by `<lower=...>` or `<upper=...>`

## Model block {.smaller .build}

```{r, echo=TRUE, eval=FALSE}
model {
  // priors
  a ~ normal(0, 10);
  b ~ normal(0, 10);
  sigma ~ normal(0, 10);
  // likelihood
  y ~ normal(a+b*x, sigma);
}
```

- specify priors: `~` distribution
- specify likelihood:  
    deterministic part `a+b*x`     
    stochastic part `y ~ normal(..., sigma)`
- alternatively:

```{r, echo=TRUE, eval=FALSE}
  for(i in 1:n){
    y[i] ~ normal(a+b*x[i], sigma);
  }
```

## Optional blocks

```{r, echo=TRUE, eval=FALSE}
functions {
}

transformed data {
}

transformed parameters {
}

generated quantities { 
}
```

## Calling the MCMC sampler from R {.build}

put data in a named list
```{r, echo=TRUE, eval=FALSE}
data = list(n=n, x=x, y=y)
```

compile the model (translates text into fast C++ code)
```{r, echo=TRUE, eval=FALSE}
stan_model = stan_model(model_code=stan_code)
```

start the MCMC sampler
```{r, echo=TRUE, eval=FALSE}
fit = sampling(stan_model, data=data) # + additional input
```
